The core transcriptome of mammalian placentas and the divergence of expression with placental shape

### 2017-03-09 manuscript/manuscript.Rnw
fix minor issues; add acknowledgements

Even though the placenta's primary function is conserved among
mammals, it varies widely in terms of shape, degree
of intimacy between fetal tissues and maternal tissues, and degree of
interdigitation of maternal and fetal tissues in different species of
mammals
\cite{Mossman.1987:vertebratefetalmembranes} %,Moffett.ea2006:Immunologyofplacentationin}
(Supplemental \cref{tab:placental_classification}). Mammals also
differ drastically in the number and size of their offspring, as well
as the length of gestation; these traits are often associated with
specific aspects of placental morphology
\cite{Capellini.ea2011:Placentationandmaternalinvestment}
(Supplemental \cref{tab:gestational_periods}). Although the morphology
of the mammalian placenta has been well characterized in a number of
species \cite{Mossman.1987:vertebratefetalmembranes}, the diversity of
the molecular environment is only beginning to be understood from a
genetic perspective \cite{Carter2012:Evolutionofplacentalfunction}.

@@

To address our questions, we 
sequenced and quantified the total RNA of mammalian placentas at term,
including \human, \bonobo (the sister
species to \textit{Pan troglodytes}), \smonkey,
\mouse, \galili, \carmeli, \dog, \cow, \elephant,
\armadillo, and \opossum.
These species represent four major branches of eutherian mammals
(Euarchontoglires, Laurasiatheria, Afrotheria, and Xenartha) as well
as marsupials. These placentas also represent a wide array of
placental morphologies as well as traits such as number and size of
offspring and length of gestation (\cref{tab:gestational_periods}). We
use these data, along with existing RNA-Seq data from \mouse
\cite{Wang.ea2011:surveyfornovelimprinted}, \human
\cite{Hughes.ea2015:Evaluatingintra-andinter-individual},
\horse
\cite{Coleman.ea2010:Structuralannotationofequine}, \sheep
\cite{Archibald.ea2010:sheepgenomereferencesequence}, and
\pig \cite{Ahanda.ea2012:Predictionofaltered3-} to: 1)
identify the non-housekeeping genes which are expressed in the
placentas of all eutherian species analyzed and therefore represent the
\textbf{core functionality of the placenta}; 2) identify genes whose
expression level changes significantly in each of the morphology
types.

@@

Fetal tissue was collected from the placenta of nine mammalian species
(\human [human], \bonobo [bonobo], \smonkey [spider monkey], \dog
[domestic dog], \cow [domestic cow], \elephant [African savanna
elephant], \opossum [gray short-tailed opossum], \galili [blind mole
rat from Upper Galilee], \carmeli [blind mole rat from northern
Israel] \cite{Nevo2001:adaptiveradiationofblindsubterraneanmole}; n=1
for all collected species). Information on specimen ID, number of
specimens, and collector is shown for each sample in Supplemental
\cref{tab:srr_numbers}. In the case of \bonobo, \cow, and \elephant,
the placenta was collected at term upon expulsion from the mother
following delivery. Placental tissue from \human was collected upon
delivery via a cesarean section. Near-term placental tissue from \dog
was collected upon the death of the mother and fetus from a car
accident. Placental tissue from \opossum was sampled at fetal stage
32, roughly one day prior to birth. Only the fetal villous portion of
the placenta was sampled at the maternal-fetal interface (i.e. villous
portion for \human, \bonobo, \dog, and \elephant; cotyledon for \cow;
diffuse placenta for \opossum). The placenta membranes were separated
from the fetal villous portion using forceps and a scapel, and the
resulting fetal villous portion was washed twice in PBS. All tissues
were collected and stored in RNAlater (Life Technologies, Carlsbad,
CA) until RNA isolation could be performed. Tissues were homogenized
in Trizol, and total RNA was isolated following the manufacturer's
instructions. RNA was further purified using RNeasy Midi Kit (Qiagen,
Valencia, CA), in conjunction with the RNase-Free DNase Set (Qiagen,
Valencia, CA), following the manufacturer’s protocol to remove any
contaminating DNA from the sample. Total RNA was then evaluated on a
Nanodrop 1000 (Thermo Scientific, Wilmington, DE) and an Agilent
Bioanalyzer 2100 (Santa Clara, CA) to confirm concentration and
quality of the isolation. RNA quality was assessed using A260/A280 and
A260/A230 absorbance values.

@@

Orthologous pairs between \human and all other species were identified
using the \ac{OMA} database
\cite{Altenhoff.ea2015:OMAorthologydatabasein}. In cases where a gene
did not have a human ortholog pair identified by \ac{OMA}, Ensembl v.80
one2one orthologs obtained from Biomart using the biomaRt R package
\cite{Durinck.ea2009:Mappingidentifiersforintegration} %
were used to select the appropriate orthologous pair. A total of
\Sexpr{combined.fpkm[,list(n=length(gene_id),
  max.fpkm=max(mean_fpkm)),by=human_name][n==14 & max.fpkm >
  10,length(human_name)]} ortholog pair sets in all 14 species were
expressed (\ac{FPKM} $\ge 10$) in at least one species. We also
obtained orthology from the perspective of each species to all other
species using biomaRt package and Ensembl v.80 database to determine
the percentage of highly expressed genes which had an orthology of
one-to-one, one-to-many, many-to-many, or novel/unknown. For future
analyses by other investigators, we have also provided the human gene
which most closely aligns to each expressed annotated
protein-generating sequence of all species analyzed using DIAMOND
\cite{Buchfink.ea2015:Fastandsensitiveprotein}.

Significant shifts in expression levels between groups of species were
calculated using Student's t test on ortholog pairs using the mean of
each species to avoid large biases towards samples with $n \gg 1$ and
considered significant if the \ac{FDR} as estimated by the Benjamini
and Hochberg procedure \cite{Benjamini1995:bh_procedure} was
$\le 0.05$. For \human and \mouse lineage specific calculations, a
type-III anova was calculated using
lme4~\cite{Bates.ea2015:FittingLinearMixed-EffectsModels} with each
species having separate random effects. Shifts in expression levels
are reported with $\log_2$ Fold-Change (FC), rounded to the nearest
integer.

@@

All of the code used to produce every analysis presented in this paper
is available at
\url{https://github.com/uiuc-cgm/placenta-viviparous.git}. All of the
sequences used are deposited in SRA, and new sequences for this work
can be obtained under the series
\href{http://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE79121}{GSE79121}
(Supplemental \cref{tab:srr_numbers}). The mean \ac{FPKM} for gene in each
species with ortholog annotations are located in the supplemental data
file all\_species\_mean\_fpkm.txt.xz and the per-sample \ac{FPKM} values are in
all\_species\_per\_sample\_fpkm.txt.xz.

@@

Supplemental \cref{fig:transcriptome_statistics} shows the RNA-Seq read and
alignment statistics for all species. For transcriptomes sequenced
with 76 bp %
%
(\Sexpr{species(star.logs.wide[input.read.length > 70 & input.read.length < 80,unique(species)])}),%
%
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,max(as.numeric(input.reads))]}
with an average of \Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,mean(`% good reads`)]}\% of reads 
mapping to the respective reference genome (\cref{tab:ref_genome_aligned_species}). 
%
For transcriptomes sequenced with 150 bp %
% The mismatch here is because we're reporting trimmed reads here, not
% full length reads
(\Sexpr{species(star.logs.wide[input.read.length > 120 & input.read.length < 160,unique(species)])}),
%
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length > 120 & input.read.length < 160,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length  > 120 & input.read.length < 160,max(as.numeric(input.reads))]}
%
with an average of \Sexpr{star.logs.wide[input.read.length > 120 & input.read.length < 160,mean(`% good reads`)]}\% %
of reads mapping to the
respective reference genome (Supplemental \cref{tab:ref_genome_aligned_species}).
%
%
For transcriptomes sequenced with 100 bp %
% The mismatch here is because we're reporting trimmed reads here, not
% full length reads
(\Sexpr{species(star.logs.wide[input.read.length > 80 & input.read.length < 120,unique(species)])}),
%
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length > 80 & input.read.length < 120,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length  > 80 & input.read.length < 120,max(as.numeric(input.reads))]}
%
with an average of \Sexpr{star.logs.wide[input.read.length > 80 & input.read.length < 120,mean(`% good reads`)]}\% %
of reads mapping to the
respective reference genome (Supplemental \cref{tab:ref_genome_aligned_species}).
%
%
For transcriptomes sequenced with 35 bp %
% The mismatch here is because we're reporting trimmed reads here, not
% full length reads
(\Sexpr{species(star.logs.wide[input.read.length < 40 ,unique(species)])}),
%
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length < 40 ,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length  < 40  ,max(as.numeric(input.reads))]}
%
with an average of %
\Sexpr{star.logs.wide[input.read.length < 40, mean(`% good reads`)]}\%
%
of reads mapping to the respective reference genome (Supplemental
\cref{tab:ref_genome_aligned_species}). Read length was not a
significant predictor of unmapped read percentage (Supplemental
\cref{fig:transcriptome_reads_vs_read_length}). %
%
Between 20\% and 40\% of all highly expressed genes in each species
were ortholog pairs with all other analyzed species, with the
remainder being mostly one-to-many or many-to-many in comparison to at
least one other species (Supplemental
\cref{tab:orthologs_in_each_species}).
%
The expression of
all genes with \ac{FPKM} $\ge 50$ is shown in Supplemental
\cref{tab:all_ateles_fusciceps_placenta,tab:all_homo_sapiens_placenta,tab:all_pan_paniscus_placenta,tab:all_mus_musculus_placenta,tab:all_spalax_galili_placenta,tab:all_spalax_carmeli_placenta,tab:all_bos_taurus_placenta,tab:all_ovis_aries_placenta,tab:all_sus_scrofa_placenta,tab:all_equus_caballus_placenta,tab:all_canis_familiaris_placenta,tab:all_loxodonta_africana_placenta,tab:all_dasypus_novemcinctus_placenta,tab:all_monodelphis_domestica_placenta}.

@@

The core placenta transcriptome contains genes which are involved in
maternal-fetal tolerance, such as the annexins, and genes responsible
for cell-cell and cell-matrix interactions, such as \ac{LGALS1}. These
genes were almost certainly present in the placenta of the most recent
common mammalian ancestor included in this study. We hypothesize that
modifications to their expression or function will compromise the
development and organization of the placenta.

@@

  \section*{Acknowledgements}
  The authors would like to specially acknowledge those groups whose
  publicly available data contributed to the completeness of this
  work.

>>>>

Even though the placenta's primary function is conserved among
mammals, the placenta varies widely in terms of shape, degree
of intimacy between fetal tissues and maternal tissues, and degree of
interdigitation of maternal and fetal tissues in different species of
mammals
\cite{Mossman.1987:vertebratefetalmembranes} %,Moffett.ea2006:Immunologyofplacentationin}
(Supplemental \cref{tab:placental_classification}). Mammals also
differ drastically in the number and size of their offspring, as well
as the length of gestation; these traits are often associated with
specific aspects of placental morphology
\cite{Capellini.ea2011:Placentationandmaternalinvestment}
(Supplemental \cref{tab:gestational_periods}). Although the morphology
of the mammalian placenta has been well characterized in a number of
species \cite{Mossman.1987:vertebratefetalmembranes}, the diversity of
the molecular environment is only beginning to be understood from a
genetic perspective \cite{Carter2012:Evolutionofplacentalfunction}.

@@

To address our questions we
sequenced and quantified the total RNA of mammalian placentas at term,
including \human [human], \bonobo [bonobo], (the sister
species to \textit{Pan troglodytes}), \smonkey [spider monkey],
\mouse [mouse], \galili [blind mole rat from Upper Galilee], \carmeli [blind mole rat from northern Israel], \dog [domestic dog], \cow [domestic cow], \elephant [African savanna elephant],
\armadillo [nine-banded armadillo], and \opossum [gray short-tailed opossum].
These species represent four major branches of eutherian mammals
(Euarchontoglires, Laurasiatheria, Afrotheria, and Xenartha) as well
as marsupials (\cref{fig:species_tree}). These placentas also represent a wide array of
placental morphologies as well as traits such as number and size of
offspring and length of gestation (Supplemental \cref{tab:gestational_periods}). We
use these data, along with existing RNA-Seq data from \mouse 
\cite{Wang.ea2011:surveyfornovelimprinted}, \human
\cite{Hughes.ea2015:Evaluatingintra-andinter-individual},
\horse [horse]
\cite{Coleman.ea2010:Structuralannotationofequine}, \sheep 
\cite{Archibald.ea2010:sheepgenomereferencesequence}, and
\pig [domestic pig] \cite{Ahanda.ea2012:Predictionofaltered3-} to: 1)
identify the non-housekeeping genes which are expressed in the
placentas of all eutherian species analyzed and therefore represent the
\textbf{core functionality of the placenta}; 2) identify genes whose
expression level changes significantly in each of the placenta morphology
types.

@@

Fetal tissue was collected from the placenta of nine mammalian species
(\human [human], \bonobo [bonobo], \smonkey [spider monkey], \dog
[domestic dog], \cow [domestic cow], \elephant [African savanna
elephant], \opossum [gray short-tailed opossum], \galili [blind mole
rat from Upper Galilee], \carmeli [blind mole rat from northern
Israel] \cite{Nevo2001:adaptiveradiationofblindsubterraneanmole}; n=1
for all collected species). Information on specimen ID, number of
specimens, and collector is shown for each sample in Supplemental
\cref{tab:srr_numbers}. In the case of \bonobo, \cow, and \elephant,
the placenta was collected at term upon expulsion from the mother
following delivery. Placental tissue from \human was collected upon
delivery via a cesarean section. Near-term placental tissue from \dog
was collected upon the death of the mother and fetus from a car
accident. Placental tissue from \opossum was sampled at fetal stage
32, roughly one day prior to birth. Only the fetal villous portion of
the placenta was sampled at the maternal-fetal interface (i.e. villous
portion for \human, \bonobo, \dog, and \elephant; cotyledon for \cow;
diffuse placenta for \opossum). The placenta membranes were separated
from the fetal villous portion by dissection, and the
resulting fetal villous portion was washed twice in PBS. All tissues
were collected and stored in RNAlater (Life Technologies, Carlsbad,
CA) until RNA isolation could be performed. Tissues were homogenized
in Trizol, and total RNA was isolated following the manufacturer's
instructions. RNA was further purified using RNeasy Midi Kit (Qiagen,
Valencia, CA), in conjunction with the RNase-Free DNase Set (Qiagen,
Valencia, CA), following the manufacturer’s protocol to remove any
contaminating DNA from the sample. Total RNA was then evaluated on a
Nanodrop 1000 (Thermo Scientific, Wilmington, DE) and an Agilent
Bioanalyzer 2100 (Santa Clara, CA) to confirm concentration and
quality of the isolation. RNA quality was assessed using A260/A280 and
A260/A230 absorbance values.

@@

Orthologous pairs between \human and all other species were identified
using the \ac{OMA} database
\cite{Altenhoff.ea2015:OMAorthologydatabasein}. In cases where a gene
did not have a human ortholog pair identified by \ac{OMA}, Ensembl v.80
one2one orthologs obtained from Biomart using the biomaRt R package
\cite{Durinck.ea2009:Mappingidentifiersforintegration} %
were used to select the appropriate orthologous pair. A total of
\Sexpr{combined.fpkm[,list(n=length(gene_id),
  max.fpkm=max(mean_fpkm)),by=human_name][n==14 & max.fpkm >
  10,length(human_name)]} ortholog pair sets in all 14 species were
expressed (\ac{FPKM} $\ge 10$) in at least one species. We also
obtained orthology from the perspective of each species to all other
species using biomaRt package and Ensembl v.80 database to determine
the percentage of highly expressed genes which had an orthology of
one-to-one, one-to-many, many-to-many, or novel/unknown. For future
analyses by other investigators, we have also provided the human gene
which most closely aligns to each expressed annotated
protein-generating sequence of all species analyzed using Diamond
\cite{Buchfink.ea2015:Fastandsensitiveprotein}.

Significant shifts in expression levels between groups of species were
calculated using Student's t test on ortholog pairs using the mean of
each species to avoid large biases towards samples with $n \gg 1$ and
considered significant if the \ac{FDR} as estimated by the Benjamini
and Hochberg procedure \cite{Benjamini1995:bh_procedure} was
$\le 0.05$. For \human and \mouse lineage specific calculations, a
type-III anova was calculated using
lme4~\cite{Bates.ea2015:FittingLinearMixed-EffectsModels} with each
species having separate random effects. Shifts in expression levels
are reported in $\log_2$ Fold-Change (FC).

@@

All of the code used to produce every analysis presented in this paper
and the manuscript itself is available at
\url{https://github.com/uiuc-cgm/placenta-viviparous.git}. All of the
sequences used are deposited in SRA, and new sequences for this work
can be obtained under the series
\href{http://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE79121}{GSE79121}
(Supplemental \cref{tab:srr_numbers}). The mean \ac{FPKM} for gene in each
species with ortholog annotations are located in the supplemental data
file all\_species\_mean\_fpkm.txt.xz and the per-sample \ac{FPKM} values are in
all\_species\_per\_sample\_fpkm.txt.xz.

@@

Supplemental \cref{fig:transcriptome_statistics} shows the RNA-Seq read and
alignment statistics for all species. For transcriptomes sequenced
with 76 bp %
%
(\Sexpr{species(star.logs.wide[input.read.length > 70 & input.read.length < 80,unique(species)])}),%
%
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,max(as.numeric(input.reads))]}
with an average of \Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,mean(`% good reads`)]}\% of reads 
mapping to the respective reference genome (\cref{tab:ref_genome_aligned_species}). 
%
For transcriptomes sequenced with 150 bp %
% The mismatch here is because we're reporting trimmed reads here, not
% full length reads
(\Sexpr{species(star.logs.wide[input.read.length > 120 & input.read.length < 160,unique(species)])}),
%
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length > 120 & input.read.length < 160,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length  > 120 & input.read.length < 160,max(as.numeric(input.reads))]}
%
with an average of \Sexpr{star.logs.wide[input.read.length > 120 & input.read.length < 160,mean(`% good reads`)]}\% %
of reads mapping to the
respective reference genome (Supplemental \cref{tab:ref_genome_aligned_species}).
%
%
For transcriptomes sequenced with 100 bp %
% The mismatch here is because we're reporting trimmed reads here, not
% full length reads
(\Sexpr{species(star.logs.wide[input.read.length > 80 & input.read.length < 120,unique(species)])}),
%
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length > 80 & input.read.length < 120,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length  > 80 & input.read.length < 120,max(as.numeric(input.reads))]}
%
with an average of \Sexpr{star.logs.wide[input.read.length > 80 & input.read.length < 120,mean(`% good reads`)]}\% %
of reads mapping to the
respective reference genome (Supplemental \cref{tab:ref_genome_aligned_species}).
%
%
For transcriptomes sequenced with 35 bp %
% The mismatch here is because we're reporting trimmed reads here, not
% full length reads
(\Sexpr{species(star.logs.wide[input.read.length < 40 ,unique(species)])}),
%
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length < 40 ,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length  < 40  ,max(as.numeric(input.reads))]}
%
with an average of %
\Sexpr{star.logs.wide[input.read.length < 40, mean(`% good reads`)]}\%
%
of reads mapping to the respective reference genome (Supplemental
\cref{tab:ref_genome_aligned_species}). Read length was not a
significant predictor of unmapped read percentage (Supplemental
\cref{fig:transcriptome_reads_vs_read_length}). %
%
Between 20\% and 37\% of all highly expressed genes in each species
were ortholog pairs with all other analyzed species, with the
remainder being mostly one-to-many or many-to-many in comparison to at
least one other species (Supplemental
\cref{tab:orthologs_in_each_species}).
%
The expression of
all genes with \ac{FPKM} $\ge 50$ is shown in Supplemental
\cref{tab:all_ateles_fusciceps_placenta,tab:all_homo_sapiens_placenta,tab:all_pan_paniscus_placenta,tab:all_mus_musculus_placenta,tab:all_spalax_galili_placenta,tab:all_spalax_carmeli_placenta,tab:all_bos_taurus_placenta,tab:all_ovis_aries_placenta,tab:all_sus_scrofa_placenta,tab:all_equus_caballus_placenta,tab:all_canis_familiaris_placenta,tab:all_loxodonta_africana_placenta,tab:all_dasypus_novemcinctus_placenta,tab:all_monodelphis_domestica_placenta}.

@@

The core placenta transcriptome contains genes which are involved in
maternal-fetal tolerance, such as the annexins, and genes responsible
for cell-cell and cell-matrix interactions, such as \ac{LGALS1}. These
genes were almost certainly present in the placenta of the most recent
common mammalian ancestor of the species included in this study. We hypothesize that
modifications to the expression or function of genes in the core placenta transcriptome
will compromise the
development and organization of the placenta.

@@

  \section*{Acknowledgements}
  The authors would like to specially acknowledge those groups whose
  publicly available data contributed to the completeness of this
  work, as well as the suggestions of the anonymous reviewers which
  improved this paper and the advice of Monica Uddin in preparing this
  manuscript.
